Actuator line modeling of vertical-axis turbines

### 2016-04-29 paper.tex
Start adding turbine geometry and CFT kinematics

Very fast solution times can be obtained with blade element momentum (BEM)
models, such as the double multiple streamtube (DMST) approached developed by
Paraschivoiu~\cite{Para1988}. However, their poor performance for high solidity
turbines \cite{Joo2015}, makes them less attractive, particular for marine
hydrokinetic applications. Their description of the flow field is very crude as
well, simply computing the momemtum contained within discretized streamtubes,
which cannot resolve any of the complicated flow structures shed by the rotor.

@@

@@

The actuator line model is based on the classical blade element theory combined
with a Navier--Stokes description of the flow field. The ALM treats turbine
blades as lines of blade elements, for which 2-D profile lift and drag
coefficients are given. For each blade element, relative flow velocity and angle
of attack are computed by adding the vectors of relative blade motion and the
local fluid velocity. The blade lift and drag forces are calculated as
\begin{equation}
    F_l = \frac{1}{2} \rho A_\mathrm{elem} C_l |\vec{U}_\mathrm{rel}|^2,
\end{equation}
and
\begin{equation}
    F_d = \frac{1}{2} \rho A_\mathrm{elem} C_d |\vec{U}_\mathrm{rel}|^2,
\end{equation}
where $\rho$ is the fluid density, $A_\mathrm{elem}$ is the blade element
planform area (span $\times$ chord), $\vec{U}_\mathrm{rel}$ is the local
relative velocity, and $C_l$ and $C_d$ are the sectional lift and drag
coefficients, linearly interpolated from a table per the local angle of attack.
The forces are then projected onto the rotor coordinate system to calculate
torque, overall drag, etc. Forces from the turbine shaft and blade support
struts are computed in a similar way. After the force on the actuator lines from
the flow is computed, it is then added to the Navier--Stokes equations as a body
force or momentum source (per unit density, assuming incompressible flow):
\begin{equation}
    \frac{\mathrm{D} \vec{u}}{\mathrm{D} t} = - \frac{1}{\rho} \nabla p + \nu
    \nabla^2 \vec{u} + F_\mathrm{turbine}.
\end{equation}

@@

    \includegraphics[width=0.9\textwidth]{thesis_alm-geometry}

@@

Troldborg~\cite{Troldborg2008} proposed that the Gaussian width should be set to
twice the local cell length $\Delta x$ in order to maintain numerical stability.

Jha et al.~\cite{Jha2014} provides guidelines for choosing a projection
width.

Schito and Zasso \cite{Schito2014} found that a projection $\epsilon$ equal to
the local mesh length was optimal.

@@

    \includegraphics[width=0.75\textwidth]{end-effects_rvat-20-deg}

>>>>

Very fast solution times can be obtained with blade element momentum (BEM)
models, such as the double multiple streamtube (DMST) approached developed by
Paraschivoiu~\cite{Para1988}. However, their poor performance for high solidity
(or chord-to-radius ratio $c/R$) turbines \cite{Joo2015}, makes them less
attractive, particular for marine hydrokinetic applications. Their description
of the flow field is very crude as well, simply computing the momemtum contained
within discretized streamtubes, which cannot resolve any of the complicated flow
structures shed by the rotor.

@@

\section{Turbines}

Two turbines were modeled: the high solidity ($c/R=0.28$) UNH Reference
Vertical-Axis Turbine (UNH-RVAT) and the medium solidity ($c/R=0.07$--$0.12$) US
Department of Energy/Sandia National Labs Reference Model 2 (RM2) cross-flow
turbine (at 1:6 scale), both of which are shown in Figure~\ref{fig:turbines}.
Both rotors are of similar geometric size ($\sim 1$ m) and are constructed from
symmetrical NACA foils mounted at mid-chord and mid-span.

Open datasets for the performance and near-wake dynamics of the UNH-RVAT and
RM2, available from \cite{Bachant2014-RVAT-baseline,Bachant2016-RVAT-Re-dep} and
\cite{Bachant2016-RM2-data}, respectively. have been acquired in the UNH towing
tank at Reynolds numbers based on turbine diameter $Re_D = 10^6$. Analysis of
these datasets showed that turbine performance and near-wake characteristics
were nearly Reynolds number independent in this
regime~\cite{Bachant2016-Energies,Bachant2016-RM2-paper}.

    \includegraphics[width=0.6\textwidth]{turbines}

    \caption{Drawings of the turbines used in this study: the UNH-RVAT (left)
    and DOE/SNL RM2 (right).}

    \label{fig:turbines}
\end{figure}

@@

The actuator line model is based on the classical blade element theory combined
with a Navier--Stokes description of the flow field. The ALM treats turbine
blades as lines of blade elements, for which 2-D profile lift and drag
coefficients are given. For each blade element, relative flow velocity and angle
of attack are computed by adding the vectors of relative blade motion and the
local fluid velocity, a diagram of which is shown in Figure~\ref{fig:vectors}.

    \includegraphics[clip, trim=1in 1.5in 1in 0.5in,
    width=0.4\textwidth]{figures/CFT-vectors_cft-vectors}

    \caption{Vector diagram of velocity and forcing on a cross-flow turbine
    blade element. Note that the free stream flow is from top to bottom, the
    blade chord is in line with the tangential velocity (i.e., zero preset
    pitch), and the drag vector is magnified by a factor of two (approximately,
    relative to the lift vector) to enhance visibility.}

    \label{fig:vectors}
\end{figure}

The kinematics of a CFT are parameterized by the tip speed ratio $\lambda =
\omega R/U_\infty$, where $\omega$ is the shaft's angular velocity, $R$ is the
maximum rotor radius, and $U_\infty$ is the free stream velocity. Using only
geometric considerations---that is, ignoring the effect of the rotor forcing and
assuming the inflow velocity to be equal to that of the free stream---the
geometric angle of attack $\alpha$ and relative velocity magnitude
$|\vec{U}_{\mathrm{rel}}|$ can be calculated as a function of the rotor azimuthal
angle $\theta$, which is plotted for various tip speed ratios in
Figure~\ref{fig:geom-alpha-urel}.

    \includegraphics[width=0.75\textwidth]{CFT-vectors_alpha_deg_urel_geom}

    \caption{Geometric angle of attack (left) and relative velocity (right)
        versus azimuthal angle at various tip speed ratios.}

    \label{fig:geom-alpha-urel}
\end{figure}

The blade lift and drag forces are calculated as
\begin{equation}
    F_l = \frac{1}{2} \rho A_\mathrm{elem} C_l |\vec{U}_\mathrm{rel}|^2,
\end{equation}
and
\begin{equation}
    F_d = \frac{1}{2} \rho A_\mathrm{elem} C_d |\vec{U}_\mathrm{rel}|^2,
\end{equation}
where $\rho$ is the fluid density, $A_\mathrm{elem}$ is the blade element
planform area (span $\times$ chord), $\vec{U}_\mathrm{rel}$ is the local
relative velocity, and $C_l$ and $C_d$ are the sectional lift and drag
coefficients, linearly interpolated from a table per the local angle of attack.
The forces are then projected onto the rotor coordinate system to calculate
torque, overall drag, etc. Forces from the turbine shaft and blade support
struts are computed in a similar way. After the force on the actuator lines from
the flow is computed, it is then added to the Navier--Stokes equations as a body
force or momentum source (per unit density, assuming incompressible flow):
\begin{equation}
    \frac{\mathrm{D} \vec{u}}{\mathrm{D} t} = - \frac{1}{\rho} \nabla p + \nu
    \nabla^2 \vec{u} + F_\mathrm{turbine}.
\end{equation}

@@

    \includegraphics[width=0.7\textwidth]{thesis_alm-geometry}

@@

Troldborg~\cite{Troldborg2008} proposed that the Gaussian width should be set to
twice the local cell length $\Delta x$ in order to maintain numerical stability.
Schito and Zasso \cite{Schito2014} found that a projection $\epsilon$ equal to
the local mesh length was optimal. Jha et al.~\cite{Jha2014} investigated the
ideal projection width for HAWT blades, recommending an equivalent elliptic
planform be constructed and used to calculate a spanwise $\epsilon$
distribution.

@@

    \includegraphics[width=0.65\textwidth]{end-effects_rvat-20-deg}
